Case-Base Neural Network: survival analysis with time-varying, higher-order interactions

### 2024-01-09 manuscript/manuscript.tex
MLwA update
fixed referencing
Updated manuscript
extended literature review

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01
%%
%% 
%%
%% $Id: elsarticle-template-num.tex 190 2020-11-23 11:12:32Z rishi $
%%
%%
\documentclass[preprint,12pt]{elsarticle}
\usepackage{hyperref}
 \usepackage{amsmath}
 \usepackage{float}
%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

@@

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \affiliation{organization={},
%% addressline={},
%% city={},
%% postcode={},
%% state={},
%% country={}}
%% \fntext[label3]{}

@@

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \affiliation[label1]{organization={},
%% addressline={},
%% city={},
%% postcode={},
%% state={},
%% country={}}
%%
%% \affiliation[label2]{organization={},
%% addressline={},
%% city={},
%% postcode={},
%% state={},
%% country={}}

\author[inst1]{Jesse Islam}

\affiliation[inst1]{organization={McGill University Department of Quantitative Life Sciences},%Department and Organization
 addressline={805 rue Sherbrooke O}, 
 city={Montréal},
 postcode={H3A 0B9}, 
 state={Quebec},
 country={Canada}}

\author[inst2]{Maxime Turgeon}
\author[inst1,inst3]{Robert Sladek}
\author[inst4]{Sahir Bhatnagar}

@@

\begin{abstract}
%% Text of abstract
Neural network-based survival methods can model data-driven covariate interactions. While these methods can provide better predictive performance than regression-based approaches, not all can model time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNNs) as a new approach that combines the case-base sampling framework with flexible neural network architectures. Using a novel sampling scheme and data augmentation to naturally account for censoring, we construct a feed-forward neural network that may take time as an input. CBNNs predict the probability of an event occurring at a given moment to estimate the hazard function. We compare the performance of CBNNs to regression and neural network-based survival methods in a simulation and three case studies using two time-dependent metrics. First, we examine performance on a simulation involving a complex baseline hazard and time-varying interactions to assess all methods, with CBNN outperforming competitors. Then, we apply all methods to three real data applications, with CBNNs outperforming the competing models in two studies and showing similar performance in the third. Our results highlight the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling of single event survival outcomes. An R package is available at https://github.com/Jesse-Islam/cbnn.
\end{abstract}

%%Research highlights
\begin{highlights}
\item Case-Base Neural Networks (CBNNs) provide a neural network methodology for survival analysis that leverages a statistical sampling technique, removing the need for complex implementations while providing data-driven flexibility in full hazard function estimation.
\item CBNNs provide an advantage over competing neural network survival models, as it naturally accounts for censoring, uses binary cross-entropy as the loss function and predicts a smooth in-time risk function. 
\item CBNNs outperform the competing models in a simulation, two studies and shows similar performance in the third.
\item Our results highlight the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling
framework for data-driven, time-varying interaction modeling of single event
survival outcomes.
\end{highlights}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
survival analysis, \sep machine learning \sep case-base \sep neural network
%% PACS codes here, in the form: \PACS code \sep code
%\PACS 0000 \sep 1111
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
%\MSC 0000 \sep 1111
\end{keyword}

@@

%% \linenumbers

@@

A common assumption in survival analysis is that the risk of the event of interest does not vary with
time (i.e. proportional hazards) \citep{hanley2009}. This simplifying assumption results in using Cox proportional
hazards models more often than smooth-in-time, accelerated failure time (AFT) models \citep{hanley2009}. This
preference causes analyses to be based on hazard ratios and relative risks rather than on survival curves and
absolute risks \citep{hanley2009}. The proportional hazards assumption in a Cox model may be incorrect in studies
where the disease pathogenesis may change over time \citep{coradini2000time}. For example, studies of breast cancer
incidence have shown that the risk of developing a tumor involves a time-varying interaction with tumor size
\citep{coradini2000time}. These time-varying covariates can be incorporated easily into AFT models \citep{royston2002flexible}.
However, these regression-based models require prior knowledge of potential
time-varying interactions and their quantitative effects.

Neural networks provide a data-driven approach to approximating interaction terms that benefit prediction. For example,
DeepSurv is a neural network-based proportional hazards model that implements the Cox partial log-likelihood
as a custom loss function \citep{katzman2018DeepSurv}. This results in a stepwise absolute risk curve that cannot
accommodate time-varying interactions. DeepSurv performs better than Cox proportional hazard models in the Rotterdam and German Breast Cancer Study Group and The Molecular Taxonomy of Breast Cancer International Consortium datasets \citep{katzman2018DeepSurv}. It has also been used to develop personalized prediction models for cancers in a number of studies \citep{ds1} \citep{ds2} \citep{ds3}. Previous work suggests we may change the loss function to address
non-proportional hazards \citep{faraggi1995neural}. In contrast, DeepHit assumes an inverse Gaussian distribution
as the baseline hazard \citep{lee2018DeepHit} and specifies each follow-up time of interest in the model. DeepHit
directly estimates survival curves, rather than deriving a hazard function \citep{lee2018DeepHit}. On real datasets,
DeepHit outperformed DeepSurv \citep{lee2018DeepHit}. We note that DeepSurv and DeepHit both use custom loss
functions \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}. In addition, DeepHit introduces a hyperparameter
weighing its two loss functions (negative log-likelihood and ranking losses) \citep{lee2018DeepHit}. Compared
to these neural network methods, regression-based approaches require prior specification of all interaction terms,
which makes it challenging to model covariate effects that change with time. The current neural network models
provide flexibility at the cost of clarity, while regression models provide clarity at the cost of flexibility.

We propose Case-Base Neural Networks (CBNNs) as a method for single event survival analysis that models
time-varying interactions and a flexible baseline hazard. CBNNs use the case-base sampling technique, which
allows probabilistic models to predict survival outcomes \citep{hanley2009}. After case-base sampling, we
implement a model using common neural network components that uses transformations of time as a feature
(covariate) to specify different baseline hazards. For example, by including splines of time as covariates,
case-base with logistic regression (CBLR) can approximate the Royston-Parmar flexible baseline hazard
\citep{royston2002flexible} \citep{hanley2009}. However, this still requires explicit specification of
time-varying interactions. CBNN can model both without extra tuning parameters.

In this paper, we describe the case-base sampling method and compare its properties to neural
network models, along with our hyperparameter selection procedure,
metrics and software implementation. Next, we explore the performance of CBNN, DeepSurv,
DeepHit, Cox regression and CBLR on simulated data and
describe their performance in three case studies. Finally, we explore
the implications of our results and contextualize them within neural network survival analysis in a single
event setting.

@@

@@

Case-base sampling is an alternative framework for survival analysis \citep{hanley2009},
which converts the total survival time into discrete person-moments. In this section, we detail how neural
networks explicitly incorporate time as a feature while adjusting for the sampling bias in case-base
sampled data and describe the metrics and hyperparameter selection procedure used to compare
CBNN with other software packages. An R package to use CBNN is available at
\url{https://github.com/Jesse-Islam/cbnn}. The entire code base to reproduce the figures
and empirical results in this paper is available at \url{https://github.com/Jesse-Islam/cbnnManuscript}.

@@

To implement case-base sampling, we divide the total survival time for each individual into discrete person-specific moments (person-moments) and treat each person-moment as a sample. This creates a \emph{base series} of
\emph{person-moments} where an event does not occur. This \emph{base series} complements the \emph{case series}, which contains
all person-moments at which the event of interest occurs.

@@

To estimate the correct hazard function, we adjust for the bias introduced when sampling a fraction of the
study base (\(\frac{b}{B}\)) by adding the term \(\log\left(\frac{B}{b} \right)\) to offset \(\log\left(\frac{b}{B} \right)\)
during the fitting process.

@@

After case-base sampling, we pass all features, including time, into any user-defined feed-forward
component (Figure \ref{fig:NNarch}). Then, we add an offset term and pass the output through a sigmoid activation function
(Figure \ref{fig:NNarch}). Since we are interested in predicting the odds of an event occurring, the sigmoid
activation function is ideal as it is the inverse of the odds and can be used to calculate the hazard. The
general form for the neural network using CBNN is:

@@

where \(T\) is a random variable representing the event time, \(X\) is the random variable for a covariate
profile, \(f_{\theta}(X, T)\) represents any feed-forward neural network architecture, \(\log\left(\frac{B}{b}\right)\)
is the offset term to adjust for the bias (\(\log\left(\frac{b}{B}\right)\)) set by case-base sampling, \(\theta\) is the
set of parameters learned by the neural network and \(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating
a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven,
while user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility
of the hazard function.

@@

{\centering \includegraphics[width=1\linewidth]{figures/Figure1.pdf}}

\caption{Methodological steps involved in CBNN. The first step, case-base sampling, is completed before training begins.
Then, we pass this sampled data through a feed-forward neural network and add an offset to adjust for the bias inherent in
case-base sampling and apply a sigmoid activation function to create a probability. Once the neural network model
completes its training, we can convert the probability output to a hazard for the survival outcome of interest.}\label{fig:NNarch}
\end{figure}

@@

We use binary cross-entropy as our loss function \citep{gulli2017}:
\begin{align*}
L(\theta)=-\frac{1}{N} \sum^{N}_{i=1} y_{i} \cdot \log(\hat{f}_{\theta}(x_{i}, t_{i}) ) + (1-y_{i} )\cdot \log(1-\hat{f}_{\theta}(x_{i}, t_{i}) ),
\end{align*} where \(\hat{f}_{\theta}(x_{i}, t_{i})\) is our estimate
for a given covariate profile and time, \(y_{i}\) is our target value specifying whether an event occurred and \(N\) represents
the number of individuals in our training set.

Backpropagation with an appropriate minimization algorithm (such as Adam, RMSPropagation, stochastic gradient
descent) is used to optimize the parameters in the model \citep{gulli2017}. For our analysis, we use Adam as implemented
in Keras \citep{gulli2017}. While the size of the \emph{case series} is fixed as the number of events; the size of the
\emph{base series} is not restricted. We use a ratio of 100:1 \emph{base series} to \emph{case series} \citep{hanley2009}.
After fitting our model, we
convert the output to a hazard. To use CBNN for predictions, we manually set the offset term $\left(\log\left(\frac{B}{b} \right)\right)$
to 0 in the new data as we already account for the sampling bias during the fitting process.

@@

To choose our method-specific hyperparameters, we use the Integrated Brier Score (IBS) \citep{graf1999}, which is based on the
Brier Score (BS) and provides a summarized assessment of performance for each model. We assess the performance of models on
a held out dataset using two metrics: 1) the Index of Prediction Accuracy (IPA) \citep{kattan2018index}; and 2) the Inverse probability
censoring weights-adjusted time-dependent area under the receiver operating characteristic curve ($AUC_{IPCW}$) \citep{auc}. The
IPA score is a metric for both discrimination and calibration, while the $AUC_{IPCW}$ provides a metric for discrimination only.

\hypertarget{bs}{%
\subsubsection{Brier Score (BS)}\label{bs}}
The BS \citep{graf1999} is defined as \begin{align}\label{eqn:bs}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(1 - \widehat{F}(t \mid X_{i})\right)^{2}\cdot I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} + \frac{\widehat{F}(t\mid X_{i})^{2}\cdot I(T_{i}>t)}{\widehat{G}(t)}\right),
\end{align} where \(\delta_{i}=1\) shows individuals who have experienced the event, \(N\) represents the number of samples
in our dataset over which we calculate \(BS(t)\), \(T_{i}\) is the survival or censoring time of an individual, \(\widehat{G}(t)=P[c>t]\) is a non-parametric estimate of the censoring distribution and \(c\) is censoring time. The BS provides a score that accounts for the
information loss because of censoring. Once we fix our \(t\) of interest, the individuals in the dataset can be divided into three
groups. Individuals who experienced the event before \(t\) are present in the first term of the equation. The second term of the
equation includes individuals who experience the event or are censored after \(t\). Those censored before \(t\) (the remaining
individuals) are accounted for by the IPCW adjustment (\(G(\cdot)\)) present on both terms.

@@

The Integrated Brier Score (IBS) is a function of the BS (\ref{eqn:bs}) \citep{graf1999}, which is defined as \begin{align*}
IBS(t)=\int_{t}^{t_{max}}BS(t)dw(t),
\end{align*} where $w(t)=\frac{t}{t_{max}}$ and $t_{max}$ is the upper bound for the survival times of interest. As the IBS is calculated for a range of follow-up-times,
it is a useful metric to track overall performance across all of follow-up time during cross-validation. We use the
IPA score to get an estimate of performance at each follow-up time for our studies.

@@

{\centering \includegraphics[width=1\linewidth]{figures/Figure2.pdf}

@@

The MM and FLC datasets are available in the \textbf{survival} package in R
\citep{survpkg}. The Prostate dataset is available as part of the \textbf{asaur} package in R \citep{asaur}.
The code for this manuscript and its analyses can be found at \url{https://github.com/Jesse-Islam/cbnnManuscript}. The software package making CBNN
easier to implement can be found at \url{https://github.com/Jesse-Islam/cbnn}.

>>>>

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01
%%
%% 
%%
%% $Id: elsarticle-template-num.tex 190 2020-11-23 11:12:32Z rishi $
%%
%%
\documentclass[preprint,12pt,authoryear]{elsarticle}
\usepackage[monochrome]{xcolor}
\usepackage[linesnumbered,ruled]{algorithm2e}
\usepackage{hyperref}
 \usepackage{amsmath}
 \usepackage{float}
 \usepackage{xcolor}

 \SetKwComment{Comment}{/* }{ */}

\usepackage{amssymb}

@@

@@

\author[inst1]{Jesse Islam}
\ead{jesse.islam@mail.mcgill.ca}
\affiliation[inst1]{organization={McGill University Department of Quantitative Life Sciences},%Department and Organization
 addressline={805 rue Sherbrooke O}, 
 city={Montréal},
 postcode={H3A 0B9}, 
 state={Quebec},
 country={Canada}}

\author[inst2]{Maxime Turgeon}
\ead{turgeon.maxime@gmail.com}
\author[inst1,inst3]{Robert Sladek}
\ead{rob.sladek@mcgill.ca}
\author[inst4]{Sahir Bhatnagar}
\ead{sahir.bhatnagar@gmail.com}

@@

\begin{abstract}
%% Text of abstract
\textcolor{red}{In the context of survival analysis, data-driven neural network-based methods have been developed to model complex covariate effects}. While these methods may provide better predictive performance than regression-based approaches, not all can model time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNNs) as a new approach that combines the case-base sampling framework with flexible neural network architectures. Using a novel sampling scheme and data augmentation to naturally account for censoring, we construct a feed-forward neural network \textcolor{red}{that includes} time as an input. CBNNs predict the probability of an event occurring at a given moment to estimate the \textcolor{red}{full} hazard function. We compare the performance of CBNNs to regression and neural network-based survival methods in a simulation and three case studies using two time-dependent metrics. First, we examine performance on a simulation involving a complex baseline hazard and time-varying interactions to assess all methods, with CBNN outperforming competitors. Then, we apply all methods to three real data applications, with CBNNs outperforming the competing models in two studies and showing similar performance in the third. Our results highlight the benefit of combining case-base sampling with deep learning to provide a simple and flexible framework \textcolor{red}{for data-driven modeling of single event survival outcomes that estimates time-varying effects and a complex baseline hazard by design.} An R package is available at \href{https://github.com/Jesse-Islam/cbnn}{https://github.com/Jesse-Islam/cbnn}.
\end{abstract}

%%Research highlights
\begin{highlights}
\item \textcolor{red}{Case-Base Neural Networks (CBNNs) estimate the full hazard function.}
\item \textcolor{red}{Naturally accounts for censoring and predicts smooth-in-time risk functions.}
\item \textcolor{red}{Uses a simple objective function, unlike competing methods.}
\item \textcolor{red}{Models time-varying effects by design, unlike competing methods.}
\item \textcolor{red}{CBNNs outperform the competing models in a simulation and two studies.}
\end{highlights}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
survival analysis \sep machine learning \sep case-base \sep neural network

\end{keyword}

@@

@@

A common assumption in survival analysis is that the risk \textcolor{red}{ratio} of the event of interest does not vary with time \textcolor{red}{which has led to the dominance of proportional hazard models\citep{cox1972regression}} \citep{hanley2009}. This simplifying assumption \textcolor{red}{explains the popularity of Cox proportional hazards models over smooth-in-time, accelerated failure time (AFT) models and results in analyses} based on hazard ratios and relative risks rather than on survival curves and absolute risks \citep{hanley2009}. The proportional hazards assumption in a Cox model may be incorrect in studies where the disease pathogenesis may change over time. \textcolor{red}{For example, age, chronic obstructive pulmonary disease, diabetes mellitus, existing heart disease, prior heart surgery, intra-aortic balloon pump, urgent and emergent surgical priority are all preoperative factors that have demonstrated time-varying associations with long-term mortality following coronary artery bypass graft surgery \citep{gao2006time}. A time-varying effect of body mass index has been associated with all-cause mortality in hypertensive patients \citep{zhu2022time} and the effect of long-term statin use on type 2 diabetes onset varies with time \citep{na2020time}. The time-varying effect of tumor size has been associated with the risk of cancer recurrence \citep{coradini2000time}. Time-varying interactions were identified in appraisals of 28 of 40 targeted and immuno-oncology therapies reported by the National Institute for Health and Care Excellence, highlighting the importance of accounting for time-varying features in cancer treatment studies \citep{salmon2023clinical}.} These time-varying covariate \textcolor{red}{effects} can be incorporated easily into AFT models; however, this requires prior knowledge of potential time-varying interactions and their quantitative effects \citep{royston2002flexible}. \textcolor{red}{In this study, we develop a method that accounts for these time-varying effects without user specification, improving risk prediction models.}

We propose Case-Base Neural Networks (CBNNs) as a data-driven method for single event survival analysis. CBNNs use the case-base sampling technique, which allows probabilistic models to predict survival outcomes \textcolor{red}{by adjusting for a sampling-specific bias with an offset term} \citep{hanley2009}. After case-base sampling, we implement a model using common neural network components that use time as a feature to estimate time-varying interactions and a flexible baseline hazard.

In this paper, \textcolor{red}{we first review state-of-the-art deep learning methods for survival analysis to position CBNN relative to related works. Second,} we describe the case-base sampling \textcolor{red}{procedure} and compare its properties to neural network models, along with our hyperparameter selection \textcolor{red}{criteria}, metrics, and software implementation. \textcolor{red}{Third}, we compare the performance of CBNN to neural network and regression-based survival methods on simulated data and describe their performance in three case studies. \textcolor{red}{Finally}, we explore the implications of our results and contextualize them within neural network survival analysis in a single event setting.

 \color{red}
\section{Related works}\label{related}
 Deep learning approaches for survival analysis can be grouped into four broad categories: models that ignore censoring; Cox partial log-likelihood; discrete survival time; and fully parametric. We wish to distinguish models, where the innovation is provided by the neural network architecture, from frameworks, which alter how the survival outcome is interpreted. Based on this distinction, any framework can use any neural network architecture.

\subsection{Survival models that ignore censoring}

Survival analysis methods handle censored data, where the event time of interest may not be known. Several methods have been developed to ignore these censored individuals during the fitting process, simplifying the survival problem. For example, \citet{zadeh2020DeepSurvnet} model survival data in a deep learning context, where the model predicts survival rate categories. In contrast, two other studies predict the probability of a bad prognosis by integrating imaging and gene expression data, using either a bilinear network \citep{wang2021gpdbn} or an attention layer with transfer learning for the imaging features and ensemble forests for gene expression \citep{jia2023dccafn}. Though these methods predict a simple binary outcome, they are unable to estimate the hazard function. These architectures can be inserted into the CBNN framework which provides the added flexibility of estimating complex baseline hazards and time-varying effects. Rather than converting time-to-event outcomes into a risk-of-prognosis score, CBNN can model the full hazard function.

\subsection{Cox partial log-likelihood} \label{coxFrame}

Several architectures using the Cox partial log-likelihood have been proposed for conventional tabular datasets. \textcolor{black}{DeepSurv is a neural network-based proportional hazards model \citep{katzman2018DeepSurv}, which has been used to develop prediction models} \textcolor{red}{for several clinical scenarios \citep{ds1} \citep{ds2} \citep{ds3} \citep{she2020development}}. \citet{ching2018cox} use gene expression data as input to a single hidden layer neural network to model cancer survival; an approach which is outperformed by an autoencoder with Cox regression model \citep{yin2022cox} and a convolutional model with residual networks \citep{huang2020deep}. In parallel, \citet{hao2021deep} assess simple Cox neural networks in a high dimensional setting against random forest and penalized regression approaches. As a more complex model for gene expression, \citet{meng2022novel} propose to first fit a generative adversarial network (GAN) to the expression data before transferring the weights to a Cox neural network. These methods provide data-agnostic architectures in clinical settings that do not take advantage of known relationships between features.

 Other approaches use the prior knowledge of existing structure between features to improve predictive performance. For example,
 \citet{chen2023pathological} transfer the embeddings of pre-trained convolutional layers on histopathological images to an autoencoder, the central layer of which is passed into a linear Cox regression to predict cervical cancer survival. Histopathological images were also used to predict lung cancer survival, using either convolutional layers \citep{zhu2017wsisa} or a convolutional transformer and Siamese network model with a modified Cox partial log-likelihood loss function that accounts for aleatoric uncertainty \citep{tang2023explainable}. \citet{kaynar2023pideel} propose a pathway-informed model integrating metabolomic data and known metabolic pathways to predict survival in patients with glioma. To assess cardiovascular risk, \citet{barbieri2022predicting} propose a sex-specific neural network model using bidirectional gated recurrent units to represent recent clinical history. These architectures focus on structure within unimodal datasets and extend the concepts proposed by data-agnostic architectures.

Other approaches integrate prior knowledge across multiple modalities to improve predictive performance. \citet{hao2022joint} use sample similarity and correlations across three genomic data modalities, then integrate them in a graph neural network to model cancer outcomes. To integrate genomic and imaging data, \citet{mobadersany2018predicting} make use of convolutional layers and \citet{li2022hfbsurv} propose a factorized bilinear network as an improvement over an unfactorized network. Finally, \citet{chen2020pathomic} integrate spatial transcriptomics as well, using both convolutional layers and graphical neural networks. Whether data-agnostic, data-specific or multi-modal, all these models make use of the Cox partial log-likelihood and assume time-invariant effects.

\subsection{Discrete-time}

Discrete-time survival models make survival predictions for specified intervals of follow-up time. For example, \textcolor{black}{DeepHit directly estimates survival probabilities and assumes an inverse Gaussian distribution as the baseline hazard \citep{lee2018DeepHit}}. The same goal is accomplished by \citet{gensheimer2019scalable} using basic hidden layers and \citet{giunchiglia2018rnn} using a recurrent neural network architecture. To estimate a piecewise hazard, \citet{kopper2022deeppamm} convert survival data into a piecewise exponential data format and propose a computationally efficient neural network. To predict cancer survival, \citet{wulczyn2020deep} incorporate prior knowledge of image motifs using convolutional layers. Discrete-time neural network models have also been used with multi-modal data. To predict long term cancer survival, \citet{vale2021long} propose a model integrating images along with transcriptomic, genomic and epigenomic data. All these methods provide a discrete estimate of survival outcomes dependent on the follow-up times of interest.

\subsection{Fully parametric framework}

 Fully parametric methods provide an estimate of the full hazard function. For example, Deep Survival Machines (DSM), uses neural networks with a mixture of distributions to fit a flexible baseline hazard \citep{dsmPaper}. This method extends to settings where the covariates are measured multiple times (i.e. time-series) \citep{nagpal2021deep}. However, the initial reports did not assess performance when time-varying effects of baseline measurements are present \citep{dsmPaper} \citep{nagpal2021deep}.

 clinical datasets often show significant time-varying effects that are relevant to disease-specific outcomes \citep{coradini2000time}\citep{gao2006time} \citep{na2020time} \citep{zhu2022time} \citep{salmon2023clinical}. Not limited to these diseases, we expect all models that predict a survival outcome may gain improved performance by accounting for non-proportional hazards. Cox partial log-likelihood based models assume the effects are time invariant \citep{cox1972regression}. CBNN accounts for the time-varying effects by design. For all the discrete-time frameworks and architectures, a selection of smaller intervals of time or more survival times of interest may result in a few or no events per interval, which potentially leads to instability in the optimization function \citep{RJ-2022-052}. In contrast, selecting larger intervals may mask nonlinear, time-varying effects in the hazard function \citep{RJ-2022-052}. CBNNs can model time-varying interactions and a complex baseline hazard function without the trade-off between long or short interval lengths. CBNNs also estimate a full hazard function for all of follow-up time, rather than discrete intervals of follow-up-time. Compared to fully parametric approaches like DSM, CBNN explicitly incorporates time into the model, providing a simple way to account for time-varying effects that improve survival prediction.

@@

\color{black}

@@

Case-base sampling is an alternate framework for survival analysis \citep{hanley2009},
which converts the total survival time into discrete \textcolor{red}{person-time coordinates} (person-moments). In this section, we detail how neural networks explicitly incorporate time as a feature while adjusting for the sampling bias in case-base sampled data\textcolor{red}{;} describe the metrics\textcolor{red}{;} and \textcolor{red}{describe the} hyperparameter selection procedure used to compare
CBNN with other methods. An R package to use CBNN is available at \url{https://github.com/Jesse-Islam/cbnn}. The entire code base to reproduce the figures and empirical results in this paper is available at \url{https://github.com/Jesse-Islam/cbnnManuscript}.

@@

To implement case-base sampling, we divide the total survival time for each individual into discrete person-moments and treat each person-moment as a sample. This creates a \emph{base series} of \emph{person-moments} where an event does not occur. This \emph{base series} complements the \emph{case series}, which contains all person-moments at which the event of interest occurs.

@@

To estimate the correct hazard function, we adjust for the bias introduced when sampling a fraction of the study base (\(\frac{b}{B}\)) by adding the term \(\log\left(\frac{B}{b} \right)\) to offset \(\log\left(\frac{b}{B} \right)\) during the fitting process. \textcolor{red}{See Algorithm \ref{alg:1} for a pseudocode implementation of the case-base sampling procedure. For a discussion of the connection between our objective function and the full data likelihood, see \cite{saarela2016case}.}

\begin{algorithm}
\color{red}
 \SetKwInOut{Input}{Input}
 \SetKwInOut{Output}{Output}
 \underline{function sampleCaseBase$(X)$}\;
 \Input{X = matrix of covariate profiles. \\ $X_{T}$ = Vector of event times for each profile. \\ $X_{Y}$ = Vector of event indicators for each profile.}
 \Output{Case-base sampled data matrix.}
 $n$ = length($X_{T}$) \Comment*[r]{Number of subjects}
 $B$ = Sum($X_{T}$) \Comment*[r]{Total person-time in study base}
 $c$ = Sum($X_{Y}$) \Comment*[r]{Number of cases}
 $b$ = 100 * c \Comment*[r]{Number of base series samples}
 \For{$i = 1, \ldots, n$}{
 Set $p_i = X_T[i]/B$\;
}
\While{$j < b$}{
 Sample $\ell$ from $\{1, \ldots, n\}$ with probabilities $p_1, \ldots, p_n$\\
 Set $baseSeries[j, ] = X[\ell, ]$\\
 Sample $baseSeries_T[j]$ from $\mathrm{Uniform}(0, X_t[\ell])$\\
}
 Set $caseSeries = X[X_{Y}==1, ]$.\\
 return(concatenate($baseSeries$,$caseSeries$))
 \caption{Pseudocode algorithm of the case-base sampling procedure.}\label{alg:1}
\end{algorithm}

\color{black}

@@

After case-base sampling, we pass all features, including time, into any user-defined feed-forward component (Figure \ref{fig:NNarch}). Then, we add an offset term and pass the output through a sigmoid activation function (Figure \ref{fig:NNarch}). Since we are interested in predicting the odds of an event occurring, the sigmoid activation function is ideal as it is the inverse of the odds and can be used to calculate the hazard. The general form for \textcolor{red}{any feed-forward} neural network \textcolor{red} {architecture} using CBNN is:

@@

\noindent where \(T\) is a random variable representing the event time, \(X\) is the random variable for a covariate profile, \(f_{\theta}(X, T)\) represents any feed-forward neural network architecture, \(\log\left(\frac{B}{b}\right)\) is the offset term to adjust for the bias (\(\log\left(\frac{b}{B}\right)\)) set by case-base sampling, \(\theta\) is the set of parameters learned by the neural network and \(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven, while user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility of the hazard function.

@@

{\centering \includegraphics[width=1\linewidth]{Figure1.pdf}}

\caption{Methodological steps involved in CBNN. The first step, case-base sampling, is completed before training begins. Then, we pass this sampled data through a feed-forward neural network\textcolor{red}{,} add an offset to adjust for the bias inherent in case-base sampling and apply a sigmoid activation function to \textcolor{red}{estimate} a probability. Once the neural network model completes its training, we can convert the probability to a hazard for the survival outcome of interest.}\label{fig:NNarch}
\end{figure}

@@

We use binary cross-entropy as our loss function \citep{gulli2017}\textcolor{red}{, from which we calculate}:
\begin{align*}
L(\theta)=-\frac{1}{N}\left( \sum^{N}_{i=1} y_{i} \cdot \log(\hat{f}_{\theta}(x_{i}, t_{i}) ) + (1-y_{i} )\cdot \log(1-\hat{f}_{\theta}(x_{i}, t_{i}) ) \right),
\end{align*} where \(\hat{f}_{\theta}(x_{i}, t_{i})\) is our estimate for a given covariate profile and time, \(y_{i}\) is our target value specifying whether an event occurred, and \(N\) represents the number of individuals in our training set.

Backpropagation with an appropriate minimization algorithm (such as Adam, RMSPropagation, stochastic gradient descent) is used to optimize the parameters in the model \citep{gulli2017}. For our analysis, we use Adam \textcolor{red}{\citep{kingma2014adam}}. While the size of the \emph{case series} is fixed as the number of events; the size of the \emph{base series} is not restricted. We use a ratio of 100:1 \emph{base series} to \emph{case series} \citep{hanley2009}. After fitting our model, we convert the output to a hazard. To use CBNN for predictions, we manually set the offset term $\left(\log\left(\frac{B}{b} \right)\right)$ to 0 in the new data as we already account for the sampling bias during the fitting process.

@@

To choose our method-specific hyperparameters, we use the Integrated Brier Score (IBS) \citep{graf1999}, which is based on the Brier Score (BS) and provides a summarized assessment of performance for each model. We assess the performance of models on a held-out dataset using two metrics: 1) the Index of Prediction Accuracy (IPA) \citep{kattan2018index}; and 2) the Inverse probability censoring weights-adjusted time-dependent area under the receiver operating characteristic curve ($AUC_{IPCW}$) \citep{auc}. The IPA score is a metric for both discrimination and calibration, while the $AUC_{IPCW}$ provides a metric for discrimination only.

\hypertarget{bs}{%
\subsubsection{Brier Score (BS)}\label{bs}}
The BS \citep{graf1999} is defined as \begin{align}\label{eqn:bs}
\resizebox{0.9\hsize}{!}{$%
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(1 - \widehat{F}(t \mid X_{i})\right)^{2}\cdot I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} + \frac{\widehat{F}(t\mid X_{i})^{2}\cdot I(T_{i}>t)}{\widehat{G}(t)}\right),
$%
}%
\end{align} where \(\delta_{i}=1\) shows individuals who have experienced the event, \(N\) represents the number of samples in our dataset over which we calculate \(BS(t)\), \(T_{i}\) is the survival or censoring time of an individual, \(\widehat{G}(t)=P[c>t]\) is a non-parametric estimate of the censoring distribution and \(c\) is censoring time. The BS provides a score that accounts for the information loss because of censoring. Once we fix our \(t\) of interest, the individuals in the dataset can be divided into three groups. Individuals who experienced the event before \(t\) are present in the first term of the equation. The second term of the equation includes individuals who experience the event or are censored after \(t\). Those censored before \(t\) (the remaining individuals) are accounted for by the IPCW adjustment (\(G(\cdot)\)) present on both terms.

@@

The Integrated Brier Score (IBS) is a function of the BS (\ref{eqn:bs}) \citep{graf1999}, which is defined as \begin{align*}
IBS(t)=\int_{t}^{t_{max}}BS(t)dw(t),
\end{align*} where $w(t)=\frac{t}{t_{max}}$ and $t_{max}$ is the upper bound for the survival times of interest. As the IBS is calculated for a range of follow-up times, it is a useful metric to track overall performance across all of follow-up time during cross-validation. We use the IPA score to get an estimate of performance at each follow-up time for our studies.

@@

{\centering \includegraphics[width=1\linewidth]{Figure2.pdf}

@@

The MM and FLC datasets are available in the \textbf{survival} package in R
\citep{survpkg}. The Prostate dataset is available as part of the \textbf{asaur} package in R \citep{asaur}.
The code for this manuscript and its analyses can be found at \url{https://github.com/Jesse-Islam/cbnnManuscript}. The software package making CBNNs easier to implement can be found at \url{https://github.com/Jesse-Islam/cbnn}.
